Describe and Explore: Using feature-based time-series clustering to understand modern intensive longitudinal data (Working Title)

### 2022-11-28 methods-and-results.Rmd
:coffin: dead code
:sparkles: feature reduction

For the feature reduction, we chose the very general \textit{principal component analysis} (PCA). While some of the more tailor-made feature selection algorithms can be more accurate and meaningful in reducing the feature dimensionality, PCAs are well-established and can broadly be applied in an automatize manner. 80% rule of thumb

we select the first `r pca_cutoff` principal components because they explain 80\% of the variance in the original `r ncol(scaled_data)` features (reducing the dimensions by `r format(round((1-pca_cutoff/ncol(scaled_data))*100, 2), nsmall=2)`\%).

@@

# Elbow method
fviz_nbclust(scaled_data, kmeans, method = "wss") +
  geom_vline(xintercept = 2, linetype = 2)+
  labs(subtitle = "Elbow method")
# Silhouette method
fviz_nbclust(scaled_data, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")
# Gap statistic [takes for ever]
# nboot = 50 to keep the function speedy.
# recommended value: nboot= 500 for your analysis.
# Use verbose = FALSE to hide computing progression.
# set.seed(123)
# fviz_nbclust(scaled_data, kmeans, nstart = 100,  method = "gap_stat", nboot = 500)+
#   labs(subtitle = "Gap statistic method")
```

feature engineering: feature extraction + feature selection

use domain knowledge to extract new variables from raw data (summarize)

options: tsfresh \citep[][]{christ2018}

\textit{k} features means $2^k – 1$ possible models

>>>>

For the feature reduction, we chose the very general \textit{principal component analysis} (PCA). Some of the more tailor-made feature selection algorithms can be more accurate in reducing the feature dimensionality and might retain feature importance information more directly, depending on the specific data structure. However, PCAs have the distinct benefit that they are well-established within the psychometric literature and can broadly be applied to a wide variety of studies in an automatize manner. As our aim is to present a general illustration that can also be adopted for more general data descriptive uses, we present the workflow using a PCA here but we encourage users to consider more specialized methods as well.

To use the principal component analysis with our extracted time series features, we first standardize all features across participants to ensure that all features are weighted equally. We then enter all `r ncol(scaled_data)` features into the analysis. The PCA will use linear transformations in such a way that the first component captures the most possible variance of the original data (e.g., by finding a vector that maximizes the sum of squared distances). The following components will then use the same method to iteratively explain the most of the remaining variance while also ensuring that the components are linearly uncorrelated. In practice, this means that the PCA decomposes the `r ncol(scaled_data)` features into `r ncol(scaled_data)` principal components but now (because of the uncorrelated linear transformations) the first few principal components will capture a majority of the variance. We can then decide how much information (i.e., variance) we are willing to sacrifice for a reduced dimensionality. A common rule of thumb is to use the principal components that jointly explain 70--90\% of the original variance (i.e., cumulative percentage explained variance). For our illustration we we select the first `r pca_cutoff` principal components that explain 80\% of the variance in the original `r ncol(scaled_data)` features (reducing the dimensionality by `r format(round((1-pca_cutoff/ncol(scaled_data))*100, 2), nsmall=2)`\%).

We would like to comment on two practical matters when using principal components --- the amount of dimensionality reduction and the interpretation of the principal components. As for the expected dimensionality reduction, given its methodology, PCAs tend to 'work better' with (highly) correlated variables. Thus, with a set of very homogeneous variables and features users will need less principal components to explain a large amount of variance, while a more diverse set of variables and features will tend to require more principal components to capture the same amount of variance. While our `r pca_cutoff` principal components is still a relatively high number of variables this is not surprising as we chose a diverse conceptualization and a diverse set of time series features. As for the interpretability, PCA allows users to extract information on the meaning of the principal components. In particular, because the principal components are linear combination of original features, users can extract the relative importance of each feature for the extracted principal components (i.e., the eigenvectors). While this can be useful in understanding the variance in the original data or help with manual feature selection, we use the PCA purely to reduce the dimensionality for the clustering step. Instead of relying on the principal components, we use the original features of interest to interpret the later extracted clusters. We particularly advocate for such an approach if the original features were chosen for their psychological meaning in understanding the time series and broader phenomenon of interest.

@@

# Elbow method
fviz_nbclust(scaled_data, kmeans, method = "wss") +
  geom_vline(xintercept = 2, linetype = 2)+
  labs(subtitle = "Elbow method") +
  theme_Publication()
# Silhouette method
fviz_nbclust(scaled_data, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method") +
  theme_Publication()
# Gap statistic [takes for ever]
# nboot = 50 to keep the function speedy.
# recommended value: nboot= 500 for your analysis.
# Use verbose = FALSE to hide computing progression.
# set.seed(123)
# fviz_nbclust(scaled_data, kmeans, nstart = 100,  method = "gap_stat", nboot = 500)+
#   labs(subtitle = "Gap statistic method")
```

### 2022-11-28 methods-and-results.tex
:sparkles: feature reduction

@@

feature engineering: feature extraction + feature selection

use domain knowledge to extract new variables from raw data (summarize)

options: tsfresh \citep[][]{christ2018}

\textit{k} features means \(2^k – 1\) possible models

>>>>

For the feature reduction, we chose the very general
\textit{principal component analysis} (PCA). Some of the more
tailor-made feature selection algorithms can be more accurate in
reducing the feature dimensionality and might retain feature importance
information more directly, depending on the specific data structure.
However, PCAs have the distinct benefit that they are well-established
within the psychometric literature and can broadly be applied to a wide
variety of studies in an automatize manner. As our aim is to present a
general illustration that can also be adopted for more general data
descriptive uses, we present the workflow using a PCA here but we
encourage users to consider more specialized methods as well.

To use the principal component analysis with our extracted time series
features, we first standardize all features across participants to
ensure that all features are weighted equally. We then enter all 72
features into the analysis. The PCA will use linear transformations in
such a way that the first component captures the most possible variance
of the original data (e.g., by finding a vector that maximizes the sum
of squared distances). The following components will then use the same
method to iteratively explain the most of the remaining variance while
also ensuring that the components are linearly uncorrelated. In
practice, this means that the PCA decomposes the 72 features into 72
principal components but now (because of the uncorrelated linear
transformations) the first few principal components will capture a
majority of the variance. We can then decide how much information (i.e.,
variance) we are willing to sacrifice for a reduced dimensionality. A
common rule of thumb is to use the principal components that jointly
explain 70--90\% of the original variance (i.e., cumulative percentage
explained variance). For our illustration we we select the first 27
principal components that explain 80\% of the variance in the original
72 features (reducing the dimensionality by 62.50\%).

We would like to comment on two practical matters when using principal
components --- the amount of dimensionality reduction and the
interpretation of the principal components. As for the expected
dimensionality reduction, given its methodology, PCAs tend to `work
better' with (highly) correlated variables. Thus, with a set of very
homogeneous variables and features users will need less principal
components to explain a large amount of variance, while a more diverse
set of variables and features will tend to require more principal
components to capture the same amount of variance. While our 27
principal components is still a relatively high number of variables this
is not surprising as we chose a diverse conceptualization and a diverse
set of time series features. As for the interpretability, PCA allows
users to extract information on the meaning of the principal components.
In particular, because the principal components are linear combination
of original features, users can extract the relative importance of each
feature for the extracted principal components (i.e., the eigenvectors).
While this can be useful in understanding the variance in the original
data or help with manual feature selection, we use the PCA purely to
reduce the dimensionality for the clustering step. Instead of relying on
the principal components, we use the original features of interest to
interpret the later extracted clusters. We particularly advocate for
such an approach if the original features were chosen for their
psychological meaning in understanding the time series and broader
phenomenon of interest.

@@
